---
title: "Portfolio Theory"
subtitle: "FIN 223 Lecture 1"
author: 
  name: "Andrew Ainsworth"
  affiliation: "University of Wollongong"
format: 
  revealjs:
    smaller: true
    embed-resources: true
    html-table-processing: none
execute: 
  eval: true
  echo: false

---

## Lecture Outline
:::{}
-   A quick refresher on return and risk measurement
-   Investor preferences and utility
    -   How do we model investor’s choices between different risky portfolios?
-   Portfolio theory
    -   What happens when investors combine different assets into portfolios?
    -   What are diversification benefits?
-   Creating optimal portfolios
    -   What portfolios do investors choose to maximize their utility?
:::

-   Reading
    - Ch 5.3, 5.5, 5.6
    - Ch. 6.1-6.4


# Let's get started {footer="false"}

## Scenario return and risk

- Expected returns $(E(R))$ are defined as:
$$E(R) = \sum_{s=1}^K p_s \times R_s$$
  - $s$ represents an  economic scenario ranging from 1, 2, 3 ... to $K$
    - The scenarios could be recession, boom or normal economic times
  - $p_s$ is the probability of a certain scenario ($s$) occuring
  - $R_s$ is the asset return if a certain scenario ($s$) occurs

- Variance is defined as:
$$\sigma^2=\sum_{s=1}^K p_s \times [R_s-E(R)]^2$$

- Standard deviation $(\sigma)$ = $\sqrt{\sigma^2}$

## Scenario return and risk

```{r}
library(gt)
library(dplyr)

scenario <- data.frame(scenario = c("Excellent", "Good", "Poor", "Crash"),
                        prob = c(0.25, 0.45, 0.25, 0.05), ret = c(0.31, 0.14, -0.0675, -0.52))

scenario_gt <- scenario %>%
  gt(id = "one") %>%
  cols_label(scenario = md("**Scenario**"), prob = md("**Probability**"), ret = md("**Return**")) %>%
  fmt_percent(columns = prob, decimals = 0) %>%
  fmt_percent(columns = ret) %>%
  cols_align_decimal(columns = c(prob, ret)) %>%
  cols_align(align="center", columns = c(prob, ret)) %>%
  cols_width(everything() ~ px(250)) %>%
  tab_options(table_body.hlines.style = "none") %>%
  opt_table_font(font = google_font(name = "Montserrat"), size = 30, color = "#0b1554") %>%
  tab_style(style = cell_borders(sides = c("t", "b"), color = "#0b1554", weight = px(2)),
    locations = cells_column_labels()) %>%
  tab_style(style = cell_borders(sides = c("b"), color = "#0b1554", weight = px(2)),
    locations = cells_body(columns = everything(), rows = length(scenario)))

scenario_gt
```

$$
\begin{align}
E(R) &= \sum_{s=1}^4 p_s \times R_s \\
&= (0.25 \times 0.31) + (0.45 \times 0.14) + (0.25 \times -0.0675) + (0.05 \times -0.52) \\
&= 0.0976 \\
&=9.76\%
\end{align}
$$

## Scenario return and risk

```{r}
scenario_gt
```

$$
\begin{align}
\sigma^2 &= \sum_{s=1}^4 p_s \times [R_s-E(R)]^2 \\
&= 0.25 \times (0.31-0.0976)^2 + 0.45 \times (0.14-0.0976)^2 + \\
  & \qquad 0.25 \times (-0.0675-0.0976)^2 + 0.05 \times (-0.52-0.0976)^2 \\
&= 0.038\\
\sigma &= \sqrt{0.038}=0.1949=19.49\%\\
\end{align}
$$

## Arithmetic and geometric averages

- Arithmetic average
  - If we assume the probability of each state is equal then $p_s=\frac{1}{K}$:
$$E(R) = \sum_{s=1}^K p_s \times R_s=\frac{1}{K}\sum_{s=1}^K R_s$$


- Geometric (time-weighted) average
  - Terminal value (TV) of the investment:
$$TV_n = (1+r_1)(1+r_2)...(1+r_n)$$
  - Geometric average return $(g)$ over $n$ time periods:
$$g=TV_n^{1/n} -1$$

## Estimating variance and standard deviation

- **Population** variance is the measured as the expected value of squared deviations
$$\sigma^2 = \frac{1}{n} \sum_{t=1}^n [R_t-\bar{R}]^2$$
- We want to work with the **sample** variance and standard deviation so need to divide by $n-1$ rather than $n$
  $$\sigma^2 = \frac{1}{n-1} \sum_{t=1}^n [R_t-\bar{R}]^2$$
  $$\sigma = \sqrt{\frac{1}{n-1} \sum_{t=1}^n [R_t-\bar{R}]^2}$$

## The normal distribution

- Investment analysis can “appear” easier if we assume asset returns follow a normal distribution
  - Symmetric asset returns $\rightarrow$ Standard deviation is a good measure of risk 
  - Symmetric asset returns $\rightarrow$ Portfolio returns will be symmetric as well
  - Only mean and standard deviation needed to estimate future scenarios
  - Pairwise correlation coefficients summarize the dependence of returns across securities

- You need to be aware that we are making an assumption when we focus only on means and standard deviation

## The normal distribution

<!--
Insert chart of normal dist using r and a plot of BHP returns histogram
-->
```{r}
library(ggplot2)
ggplot(data.frame(x = c(-1, 1)), aes(x)) + stat_function(fun = dnorm, args = list(mean = 0.08, sd = 0.25))
```

# Investor decision making

## Investor decision making

- How does an investor choose the amount to invest in risky assets?
  - We can use a utility function to rank alternative outcomes
  - Investors choose among different assets or portfolios to maximise expected utility
  - Investors can have different levels of risk aversion
  - How much do they dislike risk?
  - The utility function will lead to different portfolio choices by investors

- If there is no uncertainty then we just need to determine how much we want to consume now, and how much we want to consume later
  - Risk-free asset

- If return is not certain across all possible states of the world
  - The range of possible future cash flows will impact on wealth
  - Risky assets

# Portfolio risk and return

## Portfolio with two risky assets

- For a portfolio with two risky assets, the expected return and variance equations are:
$$E(R_p) = w_1 E(R_1) + w_2 E(R_2)$$
- Correlation coefficient between two assets:

- Need to estimate two expected return, two variances and one covariance/correlation


# How do investors choose portfolios?

## Optimal combined portfolio

- Choosing the optimal risky portfolio
  - Best combination of risky and safe assets to form portfolio
  - Optimal portfolio CAL tangent to efficient frontier 
  - Slope of CAL is Sharpe Ratio of risky portfolio
$$Sharpe\ Ratio = \frac{E(R_i)-R_f}{\sigma_i}$$
- Separation Property implies portfolio choice, separated into two tasks
  1. Determine the optimal risky portfolio that contains only risky assets
  1. Determine the allocation to risky portfolio and risk-free asset based on investor's risk aversion (A)

# Conclusion

## Conclusion

- Return and risk refresher
- Use a utility function to model investor preferences
  - This allows us to combing risk and return into a single equation 
  - Allows us to incorporate different levels of risk aversion (A)
- Portfolio theory
  - What are the risk and return of portfolios?
  - How do investors benefit from diversification?
  - What portfolios maximise investor utility?
    - Without a risk-free asset
    - With a risk-free asset

## Portfolio risk and correlation
